Abstract

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Clarify dimensionality in graphical example
In response to Reviewer 2, we clarify that all the derivations are in n-dimensions, except for the graphical representation example

> [...] Computing the parameters in the orthonormalized
(or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution and apply
an oblique rotation afterwards, at least in exploratory solutions.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution applying
Equation (24).
<!-- TODO: Check Eq. nº afterwards -->
What a practitioner should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.

@@

Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that may account for any number of dimensions.
we have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:

@@

>>>>

> [...] Computing the parameters in the orthonormalized
(or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution and apply
an oblique rotation afterwards, at least in exploratory solutions.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution applying
Equation (24).
<!-- TODO: Check Eq. nº afterwards -->
What a test practitioner should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.

@@

Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that accounts for any number of dimensions.
We have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:

@@

Also, the "Graphical Representation Example" section now begins:

> Up to now, all our derivations
have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions [...]

### 2025-03-02 output/Main_text.Rmd
Clarify dimensionality in graphical example
In response to Reviewer 2, we clarify that all the derivations are in n-dimensions, except for the graphical representation example

As plotting more than two dimensions is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
In the first case, the two latent dimensions are independent,
with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
To keep the horizontal axis invariant,
we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
it only rotates the vertical axis [@harman_modern_1976]:

@@

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution and apply
an oblique rotation afterwards, at least in exploratory solutions.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution applying
Equation \@ref(eq:dir-cos-vec-cov).
What a researcher should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.

>>>>

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
In the first case, the two latent dimensions are independent,
with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
To keep the horizontal axis invariant,
we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
it only rotates the vertical axis [@harman_modern_1976]:

@@

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution and apply
an oblique rotation afterwards, at least in exploratory solutions.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution applying
Equation \@ref(eq:dir-cos-vec-cov).
What a test practitioner should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.
